### 2015-10-13 locking.tex
add discussion of second example

There are two possible interpretations.
\begin{enumerate}
	\item When a task requires a shared resource, which implies the beginning of a new segment, its lock request is processed only when its new segment becomes eligible for execution (as determined by the period enforcer algorithm). As a consequence, a task may find a required shared resource unavailable when its new segment becomes eligible for execution even though it was available when the prior segment finished.
	\item Alternatively, when a task requires a shared resource, its request may be processed \emph{immediately}. As a consequence, a shared resource may be locked by a task that is currently ineligible to execute.
\end{enumerate}
We believe that the first interpretation is the more natural one. However, for the sake of completeness, we show that either interpretation leads to potentially unbounded worst-case response times.

\subsection{Alternative 1: Locking Takes Effect at Segment Eligibility Time}
Consider a simple task set consisting of two tasks on two processors that share one resource. Task $\tau_1$, on processor~1, has a total execution cost of $C_1 = 4$ and a period and deadline of $T_1 = D_i = 8$. After one time unit of execution, jobs of $\tau_1$ require the shared resource for two time units. $\tau_1$ thus consists of two segments with cost $C_{1,1} = 1$ and $C_{1,2} = 3$. Task $\tau_2$, on processor~2, has the same overall WCET ($C_2 = 4$), a slightly shorter period ($T_2 = D_2 = 7$), and requires the shared resource for one time unit after \emph{two} time units of execution ($C_{2,1} = 2$ and $C_{2,2} = 2$).

@@

Since $\tau_2$'s first job requests the shared resource only after two time units of execution, it is blocked by $\tau_1$'s critical section, which commenced at time $1$. At time $3$, $\tau_1$ releases the shared resource and $\tau_2$ consequently resumes (i.e., $a^2_{2,1} = 3$). According to the period enforcer rules~\cite{Raj:suspension1991}, the second segment is immediately eligible because
\begin{align*}
	ET_{2,1}^2 & = \max\left(ET_{2,0}^2 + T_2,\ \mathit{busy}(\tau_2, a^2_{2,1})\right) =\max(-T_2 + T_2,\ 3) = 3.
\end{align*}
(Recall that $ET_{2,0}^2 = -T_2$, and interpret $\mathit{busy}(\tau_2, a^2_{2,1})$ with respect to $\tau_2$'s processor.)

@@

Next, we show that the alternative interpretation of when a lock request takes effect also leads to unbounded response time growth.

\subsection{Alternative 2: Locking Takes Effect Immediately}

>>>>

There are two possible interpretations.
\begin{enumerate}
	\item When a task requires a shared resource, which implies the beginning of a new segment, its lock request is processed \emph{only when its new segment would be eligible for execution}, as determined by the period enforcer algorithm. 
	\item Alternatively, when a task requires a shared resource, its request may be processed \emph{immediately}.
\end{enumerate}

As a consequence of the first rule, a task may find a required shared resource unavailable when its new segment becomes eligible for execution even though it was available when the prior segment finished.  As a consequence of the second rule, a shared resource may be locked by a task that cannot use it for the time being because it is still ineligible to execute.

We believe that the first interpretation is the more natural one, as it does not make much sense to award resources to tasks that cannot yet use them. However, for the sake of completeness, we show that either interpretation leads to potentially unbounded worst-case response times.

\subsection{Alternative 1: Locking Takes Effect at Segment Eligibility Time}
First, assume that the processing of lock requests is delayed until the point when a resuming segment would no longer be subject to any delay due to period enforcement). We show that this interpretation leads to a deadline miss in a task set that would otherwise be trivially schedulable.

To this end, consider a simple task set consisting of two tasks on two processors that share one resource. Task $\tau_1$, on processor~1, has a total execution cost of $C_1 = 4$ and a period and deadline of $T_1 = D_i = 8$. After one time unit of execution, jobs of $\tau_1$ require the shared resource for two time units. $\tau_1$ thus consists of two segments with cost $C_{1,1} = 1$ and $C_{1,2} = 3$. Task $\tau_2$, on processor~2, has the same overall WCET ($C_2 = 4$), a slightly shorter period ($T_2 = D_2 = 7$), and requires the shared resource for one time unit after \emph{two} time units of execution ($C_{2,1} = 2$ and $C_{2,2} = 2$).

@@

Since $\tau_2$'s first job requests the shared resource only after two time units of execution, it is blocked by $\tau_1$'s critical section, which commenced at time $1$. At time $3$, $\tau_1$ releases the shared resource and $\tau_2$ consequently resumes (i.e., $a^2_{2,1} = 3$). According to the period enforcer rules~\cite{Raj:suspension1991}, the second segment is immediately eligible because, according to Equation~\ref{eq:ET-def},
\begin{align*}
	ET_{2,1}^2 & = \max\left(ET_{2,0}^2 + T_2,\ \mathit{busy}(\tau_2, a^2_{2,1})\right) =\max(-T_2 + T_2,\ 3) = 3.
\end{align*}
(Recall that $ET_{2,0}^2 = -T_2$, and interpret $\mathit{busy}(\tau_2, a^2_{2,1})$ with respect to $\tau_2$'s processor.)

@@

Next, we show that the alternative interpretation of when a lock request takes effect also leads to unbounded response-time growth.

\subsection{Alternative 2: Locking Takes Effect Immediately}
From now on, assume that all lock requests are processed immediately when they are made, even if this causes the shared resource to be locked by a task that is not yet eligible to execute according to  the rules of the period enforcer algorithm. We construct an example that demonstrates unbounded response-time growth.

To this end, consider two identical tasks hosted on two processors. Task $\tau_1$ is hosted on processor~1; $\tau_2$ is hosted on processor~2. Both tasks have a period of $T_1 = T_2 = D_1 = D_2 = 8$ and a WCET of $C_1 = C_2 = 4$. They both access a single shared resource for two time units each per job. Both tasks request the shared resource after executing for \emph{at most} one time unit. Both tasks thus have two segments each with parameters $C_{1,1} = C_{2,1} = 1$ and $C_{1,2} = C_{2,2} = 3$. The example exploits that jobs may require \emph{less} service than their specified worst-case costs.

Figure~\ref{xxx} shows an example schedule assuming periodic job arrivals and that, for an arbitrary positive $\epsilon <1$,
\begin{itemize}
	\item the first segment of the first job of $\tau_2$ executes for only $1-\epsilon$ time units,
	\item the first segment of the second job of $\tau_1$ executes for only $1-\epsilon$ time units,
	\item the first segment of the third job  of $\tau_2$ executes for only $1-\epsilon$ time units, and that
	\item all other segments must execute for their specified worst-case costs.
\end{itemize}

At time $1-\epsilon$, the first job of $\tau_2$ acquires the shared resource because $\tau_1$ does not issue its request until time $1$. Consequently, $\tau_1$ is blocked until time $a^2_{1,1} = 3$, and we have
\begin{align*}
	ET_{1,1}^2 & = \max\left(ET_{1,0}^2 + T_1,\ \mathit{busy}(\tau_1, a^2_{1,1})\right) =\max(-T_1 + T_1,\ 3) = 3 
\\ \intertext{and}
	ET_{2,1}^2 & = \max\left(ET_{2,0}^2 + T_2,\ \mathit{busy}(\tau_2, a^2_{2,1})\right) =\max(-T_2 + T_2,\ 1 - \epsilon) = 1 - \epsilon.
\end{align*}

The roles of the second jobs of both tasks are reversed: since the second job of $\tau_1$ locks the shared resource already at time $9-\epsilon$, $\tau_2$ is blocked when it attempts to lock the resource at time $9$. However, according to the rules of the period enforcer algorithm, the second segment of the second job of $\tau_1$ is not actually eligible to execute before time $11$ since
\begin{align*}
	ET_{1,2}^2 & = \max\left(ET_{1,1}^2 + T_1,\ \mathit{busy}(\tau_1, a^2_{1,2})\right) =\max(3 + 8,\ 8) = 11.
\end{align*}
Consequently, even though the lock is granted to $\tau_1$ already  at time $9-\epsilon$, the critical section is executed only starting at time $11$, and $\tau_2$ is thus delayed until time $13$. At time $13$, $\tau_2$ is immediately eligible to execute since
\begin{align*}
	ET_{2,2}^2 & = \max\left(ET_{2,1}^2 + T_2,\ \mathit{busy}(\tau_2, a^2_{2,2})\right) =\max(1-\epsilon + 8,\ 13) = 13.
\end{align*}

The third jobs of both tasks are released at time $16$. The roles are swapped again: $\tau_2$ acquires the lock before $\tau_1$ can issue its request because $\tau_2$'s first segment requires only $1-\epsilon$ time units of service. However, according to the period enforcer algorithm's eligibility criterium, $\tau_2$ cannot actually continue its execution before time $21$ since
\begin{align*}
	ET_{2,3}^2 & = \max\left(ET_{2,2}^2 + T_2,\ \mathit{busy}(\tau_2, a^2_{2,3})\right) =\max(13 + 8,\ 17-\epsilon) = 21.
\end{align*}
This, however, means that $\tau_1$ cannot use the shared resource before time $23$, which leaves insufficient time to complete the -segment before $\tau_1$'s deadline at time $24$. However, as before,  the response-time analyses for both the MPCP~\cite{Br:13,LNR:09} and the   FMLP~\cite{Br:13} predict a worst-case response time of $6$ for both tasks (i.e., four time units of execution, and at most two time units of blocking).

\subsection{Discussion}

While it is intuitively appealing to combine period enforcement with suspension-based locking protocols, we observe that this causes non-trivial difficulties. In particular, our examples show that the addition of the period enforcer algorithm invalidates all existing blocking analyses. They also suggest that devising a correct blocking analysis would be a substantial challenge due to the demonstrated feedback cycle between the period enforcer rules and blocking durations. Worse, such an analysis carries the risk of being an exercise in futility, as the examples show that substantial response time increases are possible---however, at which point is the period enforcer's sole benefit, namely reduced interference on lower-priority tasks, outweighed by its cost, namely the (possibly large) increases in response times of lock-using higher-priority tasks?

Fundamentally, the period enforcer algorithm's design implicitly rests on the assumption that a segment \emph{can} execute as soon as it is eligible to do so. In the presence of locks, however, this assumption is invalidated. As demonstrated, the result can be a successive growth of self-suspension times until a deadline is missed.  In a sense, the period enforcer algorithm, at least as defined in the existing literature, is therefore incompatible with suspension-based locking protocols.

### 2015-10-13 untitled.tex
add discussion of second example

Let $ET_{2,j}^1$ and $ET_{2,j}^2$ denote the eligibility times of the first and second segment, respectively, of the $j^{\mathrm{th}}$ job of $\tau_2$. Further, let $a^1_{2,j}$ and $a^2_{2,j}$ denote the arrival times\footnote{A segment \emph{arrives} when it becomes available for execution. The first segment arrives immediately when the job is released; the segment arrives when the job resumes from its self-suspension.} of the job's first and second segment, respectively.  Finally, let $\mathit{busy}(\tau_2, t')$ denote the last time that a level-2 busy interval began on or prior to time $t'$. According to Section 3.1 in \cite{Raj:suspension1991}, the period enforcer algorithm then defines the segment eligibility times as
\begin{align*}
	ET_{2,j}^1 & = \max\left(ET_{2,j-1}^1 + T_2,\ \mathit{busy}(\tau_2, a^1_{2,j})\right)  \text{ and }
\\
	ET_{2,j}^2 & = \max\left(ET_{2,j-1}^2 + T_2,\ \mathit{busy}(\tau_2, a^2_{2,j}) \right),
\end{align*}
where $ET_{2,0}^1 = ET_{2,0}^2 = -T_2$.

Figure \ref{fig:example} shows the resulting schedule assuming a periodic release pattern. The first job of task $\tau_2$ (which arrives at time $a^1_{2,1} =  0$) is executed as if there is no period enforcement since the definition $ET_{2,0}^1 = ET_{2,0}^2 = -T_2$ ensures that both segments are immediately eligible. Note that the first segment of $\tau_2$'s first job is delayed due to interference from $\tau_1$. As a result, the second segment of $\tau_2$'s first job does not resume until time $a^2_{2,1} = 9$. Thus, we have
\begin{align*}
	ET_{2,1}^1 & = \max\left(-T_2 + T_2,\ \mathit{busy}(\tau_2, 0)\right) = 0  \text{ and }
\\
	ET_{2,1}^2 & = \max\left(-T_2 + T_2,\ \mathit{busy}(\tau_2, 9\right) ) = 9.
\end{align*}

>>>>

A segment's eligibility time is determined according to the following rule. Let $ET_{i,j}^k$ denote the eligibility time of the $k$\xth segment of the $j$\xth job of task $\tau_i$. Further, let $a^k_{i,j}$ denote the segment's arrival time.\footnote{A segment \emph{arrives} when it becomes available for execution. The first segment arrives immediately when the job is released; the segment arrives when the job resumes from its self-suspension.} Finally, let $\mathit{busy}(\tau_i, t')$ denote the last time that a level-$i$ busy interval began on or prior to time $t'$. According to Section 3.1 in \cite{Raj:suspension1991}, the period enforcer algorithm defines the segment eligibility time of the $k$\xth segment as
\begin{align}\label{eq:ET-def}
	ET_{i,j}^k & = \max\left(ET_{i,j-1}^k + T_i,\ \mathit{busy}(\tau_i, a^k_{i,j})\right),
\end{align}
where $ET_{i,0}^k = -T_i$.

Assuming this definition, Figure \ref{fig:example} shows the resulting schedule for a periodic release pattern. The first job of task $\tau_2$ (which arrives at time $a^1_{2,1} =  0$) is executed as if there is no period enforcement since the definition $ET_{2,0}^1 = ET_{2,0}^2 = -T_2$ ensures that both segments are immediately eligible. Note that the first segment of $\tau_2$'s first job is delayed due to interference from $\tau_1$. As a result, the second segment of $\tau_2$'s first job does not resume until time $a^2_{2,1} = 9$. Thus, we have
\begin{align*}
	ET_{2,1}^1 & = \max\left(-T_2 + T_2,\ \mathit{busy}(\tau_2, 0)\right) = 0  \text{ and }
\\
	ET_{2,1}^2 & = \max\left(-T_2 + T_2,\ \mathit{busy}(\tau_2, 9\right) ) = 9.
\end{align*}
